LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-06-13 article/main.tex
final draft

\textbf{Large language models (LLMs)} have are a commonplace research topic
today.
Because language models such as BERT~\cite{devlin2018bert} or
GPT~\cite{gpt-2018,gpt2-2019,gpt3-2020} constantly advance the state of the art
on many natural language processing tasks, it has become natural to want to
evaluate the performance of these generalist models on ever more specialized
tasks.
We know from multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} that large language
models (especially the more popular ones) are good at following instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended to entire articles or corpuses of text because
these are also the object of linguistic style and
stylistics~\cite{lugea2023stylistics}.
The task of transfering text style is an old preocupation for the natural
language processing and computational linguistics communities which has picked
up interest again in the context of the advancements in the field of deep
learning and particularly since the inception of the transformer architecture.

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
More often than not, these values are easy to compute while the methods used for
obtaining them are usually idempotent.
This makes them very attractive because of the dual purpose that they can serve.
On the one hand, they can be used to instruct the generation of text (``write
no more than 10 sentences'') while on the other they can be used to analyse text
and, therefore, verify generated text.

Some HLFs statistically describe text in terms of its basic syntactical and
morphological components.
Knowing this we could follow Noam Chomsky's ideas~\cite{chomsky2002syntactic}
which imply that the style of a text is influenced by its underlying syntactic
structures.
Traditionally, linguistic style was thought to be the result of morphological
and syntactic arrangements~\cite{lugea2023stylistics}.
More recent schools of thought on linguistic style find semantics and pragmatics
to be inextricably linked to author style~\cite{verma2019lexical}, but argue
that a great deal can still be inferred from lexicon, syntax and a `surface'
level which contains basic HLFs such as the number of words in a sentence.

For example, a high number of exclamations in a text is often a revealing
signal of an inflamatory news writing style, of spam or of alert narration or
dialogue.
Similar observations have been made about other characteristics of written text
that might be computed as HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

These apparently disparate strands of thought lead us to the two research
questions we'd like to answer in this paper.

@@

Linguistic style, then, is the sum of identifiable language choices manifest in
a text made from the language system by the text
producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
The ample literature on the subject seems to agree that style depends on context
and author choice vis-a-vis a communication
goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
This work grounds our choice of handcrafted linguistic features related to
linguistic style.

@@

Finally, this paper documents a form of controllable text generation
(CTG)~\cite{zhang-ctg-2022}.
Hightened recent interest in CTG and especially in benchmarking LLMs in the
context of CTG~\cite{chen2024benchmarking} is particularly relevant for this
paper.
Such research alleviates the burden of proof regarding the level to which
LLMs respond well to diversified instructions for controlled text generation in
general.
Compared to the existing work, our prompts are guided by HLFs.
Furthermore, we use those same HLFs to evaluate the performance of the LLM.

@@

Our experiment consists of analysing the behavior of an LLM when asked to
generate new text based on a given input text and some instructions derived from
HLF statistics computed on a corpus containing multiple texts.
The experiment relies on choosing LLMs to evaluate and finding datasets from
which we can derive corpuses.
Then there's the matter of selecting the HLFs for our experiment.
Let us outline the experiment process before going into further detail.

The first task is to establish a baseline.
We do so by instructing the target LLM to generate text retaining the meaning of
the input text without any further instruction related to style or the
linguistic features of the generated text.
We repeat the text generation step 10 times using the same prompt and the same
input text.

For each text generation, we compute the following statistics for each of our
chosen HLFs:

\begin{itemize}
    \item \textbf{min}: the minimum value of the HLF on the generated text;
    \item \textbf{max}: the maximum value of the HLF on the generated text;
    \item \textbf{avg}: the mean value of the HLF on the generated text;
\end{itemize}

A second baseline is obtained by following the same process and same input text,
but with a new prompt that builds upon the first one by adding the instruction
to use example text from our corpus of choice.
The examples are three texts chosen randomly once at the beginning of the
experiment from the used corpus.
We use the same example texts for all evaluated LLMs.

We choose three texts to not overrun the context window of the LLM.
This hints at the main drawback for our approach: it uses up the context window
with system instructions.
While this isn't an issue with all models, some only allow a context window of
around 8000 tokens so we should be mindful.

We construct the second baseline by conducting 10 trials using the same
prompt and input text and otherwise in the same manner in which we created the
first baseline.

The next step is to devise HLF prompts derived from both baseline prompts.
The HLF prompts contain additional instructions to generate text that would
comply with certain HLF values.
These HLF values are obtained by computing the statistics outlined above on
the entirety of the target benchmark dataset.
The HLF prompts are constructed from templates that take into account the HLF
values.
The prompt templates part of the experiment's source code.
The instructions concerning all choseon HLFs are given concomitently within the
same prompt.
In a similar manner to how we generated the baselines, we run ten trials for
each case (with or without example articles from the target benchmark dataset).

We compare the first baseline with the results obtained by using the first
modified HLF prompt to understand whether we can influence LLM text generation
using HLFs.
Our reasoning is that if the LLM were influenced by the HLF prompt, the output
would show different HLF values compared to the baseline.

We now turn our attention to the second baseline.
Given that the LLM now also has example articles, it can freely immitate those
articles when generating the text.
If there is a consistent difference over the ten trials between generating text
using the baseline or using the HLF prompt, this difference can only be a result
of the LLM complying to HLF instructions.
This difference is expressed by the change of value for each HLF between the
baseline and HLF enhanced text generation.

Because we provide the LLM with examples from the corpus, it is more likely to
intuit the linguistic features of the corpus.
To understand whether HLF instructions actually make a difference, we perform
the experiment using the second type of prompt using input text both from within
and from outside the corpus.

@@

For our experiment we will use state of the art large language models.
We used the Chatbot Arena~\cite{chiang2024chatbot} leaderboard for constructing
our selection.
In the interest of covering more ground, we chose five top performers in the
Chatbot Arena from different vendors.
The final selection is available in Table~\ref{table-llm}. The identifier

@@

The ID of the LLM is used as the reference throughout figures and tables.

@@

In this experiment we create specific instructions for each HLF.
The resulting instructions can be catalogued on the concrete-abstract spectrum.
For example, `write 100 words' is a more concrete instruction than `write as if
you were in junior highschool'~\cite{kincaid1975derivation}.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category for a grand total of 10 HLFs.
Table~\ref{table-hlf} showcases our selection of features.

@@

Each of these features is well documented~\cite{lftk-2023}, barring the need for
an extended description here.
The LFTK ID is used throughout tables and figures to refer to a specific HLF.

\subsection{DataSet Selection}\label{ds-selection}

The experiment design rests upon measuring HLFs on corpuses of texts and
computing statistics.
LLMs also make use of examples extracted from a corpus to guide text generation.
We obtain a corpus from a dataset.
Our final selection of datasets is listed in Table~\ref{table-ds}.

@@

The Yelp review data set provides a diverse assortment of reviews from online
users.
The authors employ a casual writing style, very similar to what a average
writing style of an average person would be like.
The dataset as a whole is useful for many tasks.
The corpus we construct based on this dataset contains the first 1000 reviews
from its test split.

CNN/DailyMail contains stories and news with diverse styles.
The style varies from one author to another, but also based on the publishing
entity.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative choice.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

@@

This experiment presupposes instructing various large language models to reword
an input text so that it conforms to a bunch of handcrafted linguistic features.
The input text is provided via a user prompt.

We provide two input texts that are external to our corpuses, one input text per
corpus.
The input texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.
These user prompts are used when we're not using an input text from the
corpus itself.

For the CNN/DailyMail dataset, the user prompt from outside the dataset is an
extract from the classic text \textit{The Life and Opinions of Tristram Shandy,
    Gentleman}, by Laurence Sterne~\cite{sterne2003life}.
This text is far from a news article in terms of both style and exhibited
linguistic features from the CNN/DailyMail dataset contents.
We see a text that is similarly divorced from the style and linguistic
features of most Yelp reviews in Oscar Wilde's \textit{Sonnet to
    Liberty}~\cite{wilde1909poems}.
Having input texts that are vastly different from a dataset should inform an LLM
to generate a baseline that does not conform to the linguistic features of the
dataset.

We use Jinja templates to compile the text generation instructions for the LLM.
Both the model prompts and the HLF instructions are constructed using these
templates.
The Jinja templates are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{templates}}
folder of the experiment's source code.

There are two main text generation instruction templates: one without example
content~(called `\texttt{prompt\_1}') and one with examples from the
corpus~(`\texttt{prompt\_2}').
Each of the two instruction templates allow the presence of HLF instructions.

@@

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
All our graphs show the HLF on the vertical axis, while the horizontal axis is
reserved for the trial number.
HLF values for the text generated using baseline prompts are plotted as a
continuous grey line, while the values obtained by using HLF enhanced prompts
are plotted using a continuous grenat line.
The minimum and maximum HLF on the benchmark dataset are plotted using dotted
black lines.
A continuous black line designates the average HLF on the corpus.
We sometimes refer to this average as the \textit{target} value because the LLM
is primarily instructed to conform to this value.

Result tables highlight the difference between plots obtained using a
baseline prompt versus the ones obtained using HLF-enhanced prompts.
Firstly, we care about the difference between each plot (baseline and HLF) and
the target (the corpus average).
Secondly, we measure how different baselines are from HLF-enhanced plots.

@@

To determine which of the two results is closer to the target, we compute the
area between each result and the target.
Another way to measure the geometric closeness to the target is by computing the
Euclidian norm of the vectorial difference between result and target.
Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
instruction result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
Finally, we define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$.

@@

To save space, we illustrate significant differences between the baseline and
the HLF instruction results in \textbf{bold} typeface.
The \underline{underlined} values represent situations when the HLF-enhanced
prompt lead to results that are significantly closer to the corpus average than
the baseline.
\textit{Italicized} values represent situations when the baseline was
significantly closer to the target than the HLF-enhanced result.

Lastly, it should be noted that each HLF has an independent value range.
The number of total words ranges from small numbers into the hundreds or more,
while the corrected verb variation is always a small number.
The area and the Euclidian norm are calculated using the raw values of each HLF.
That makes the values in each table column incomparable to the values in another
table column.
Because we don't show the measure of the difference between the baseline and the
result, but the relative closeness to a target based on that measure, comparing
values from the same column is also ill-advised.

In the next sections we only present the most relevant results of the
experiment.

@@

The results on the corpus containing CNN articles showcased in
Table~\ref{table-prompt-1-cnn-dailymail} shows that most LLMs take into account
instructions concerning the linguistic features of their output.

@@

The second type of instructional prompt we feed to the LLM contains three
examples from the corpus.
The hypothesis is that this will influence the baseline result to be closer to
the corpus average.

In this case, it is interesting to see how the LLM generates text based on input
text from both within the corpus that contains the example articles and from
outside the corpus.

In Table~\ref{table-prompt-2-cnn-dailymail} we see the results when an input from
outside the corpus is used.
The outlook is similar for what we saw in the first experiment with GPT
being the least responsive to HLF instructions.

@@

GPT's results would be good if the baseline were very close to the target value.
As we can see in Figure~\ref{fig-p2-gpt-twords}, this is not always the case.
We also notice that Command-R+(fig.~\ref{fig-p2-command-r-simpttr}) and Gemini
(fig.~\ref{fig-p2-gemini-a-kup-pw}) both registered significantly negative
results, in contrast to their performance in our first scenario.

@@

The results obtained in our first scenario are largely confirmed by our second
scenario.
So now we turn to see how using a text from within the corpus affects the
performance of the large language models.
The hypothesis is that if the text is `familiar', both the baseline and the HLF
instruction prompt will produce results that are much closer to the target
value.
The expectation is to see less significant results.

@@

The results shown in Table~\ref{table-prompt-2-ifd-cnn-dailymail} contradict our
expectations outright.
The most surprising behaviour is exhibited in the context of more concrete HLFs
such as the total number of words or the number of unique adjectives.
Claude3 shows regressive behaviour for the number of unique adjectives in
Figure~\ref{fig-p2-ifd-claude3-nuadj}, while Gemini constructs texts using a
very different amount of words in Figure~\ref{fig-p2-ifd-gemini-twords}.
Less surprisingly, Llama3 fails to generate text that conforms to the correct
Kuperman age of acquisition per word in Figure~\ref{fig-p2-ifd-llama3-a-kup-pw}.

@@

The results in this last scenario show that whether LLMs take examples listed in
the prompt into account or not with respect to conforming to certain linguistic
features is debatable, providing them with specific instructions derived from
HLFs makes a significant impact.

@@

In terms of our first question --- whether LLMs are able to follow instructions
concerning handcrafted linguistic features --- the response is in our opinion
affirmative.
Certainly, not all LLMs are created equal in this regard.
It is debatable whether this experiment is well designed because of the choice
to derive target HLF values for the LLM from the corpus.
This is evident in the results obtained for the Yelp dataset.
While these results also show a significant influence when using HLFs in the
prompt, the influence is mostly negative.
This can be attributed to the choice of input text in relation to the corpus,
the wording in the instruction prompt or to each model's specific training.

This leads us to our second question: how much do handcrafted linguistic feature
instructions affect the way LLMs write text.
The results in this regards show us a much less conclusive picture.

Firstly, we cannot assume that the LLM will follow HLF instructions the way we
intend.
Whether we obtain the expected outcome or not depends largely on the prompt,
the chosen HLFs, the model, the target values for the HLFs, a.s.o
We do, however, notice a few patterns.
The more abstract the instruction, the less it seems to make a positve impact.
More importantly, by inspecting the result data, a more abstract instruction
seems more likely to produce a negative impact on the desired outcome.

Secondly, the choice of input text in relation to the target HLF values seems
very important.
Choosing to reword a poem so that it has the same reading ease as a user review
seems like a setup for poor results.
This leads us to intuit that LLMs are clearly able to use HLF instructions to
generate text differently, but there is a limit to their `creativity'.

Finally, the choice of HLFs is very important.
The text generated by an LLM is less likely to conform to HLFs that result in
more abstract instructions.
Furthermore, the combination of chosen HLFs is important.
Choosing HLFs that measure the uniqueness of a linguistic element (words,
structures, a.s.o) limits the LLMs ability to conform to other HLFs (for example
total number of sentences or reading ease).
This is highlighted in the experiment specifically in the tension between the
age of acquisition and the adjective or verb variation.
Predictably, the more concrete instructions take precedence over the more
abstract ones.

@@

We started out by trying to understand whether it is possible to conform the
output of a large language model to exhibit specific linguistic features.
To our surprise, it turns out that instructing language models to write text
that conforms to a selected assortment of linguistic features almost always
yields significant changes in how the text is being generated.

This doesn't mean that the results are always satisfying.
What the model writes about and the chosen target values for the linguistic
features are major factors in how successful the overall experience is.
Unfortunately, at this stage, finding out good target values and good writing
instructions is a question of trial and error.

Other experiment designs might have been more opportune in examining the
questions layed out in the introduction.
For example, using set target values for HLFs (without deriving them from a
corpus of text) and choosing input text where the HLFs vary greatly would have
been an alternate design.
Following through on those approaches is one of the possible expansions on the
current work.

Another way of expanding on the current work is finding the settings in which
large language models are most responsive to linguistic features.
The current experiment was inconclusive in this regard and a legitimate further
question is whether settings in which text generated by language models
consistently conforms to certain linguistic features actually exist.

Lastly, a simpler question that is left unanswered relates to the number of
linguistic features that were used.
What if we would have measured models' behaviour for single linguistic features
independently?

>>>>

\textbf{Large language models (LLMs)} have become a commonplace research topic
today.
Because language models such as BERT~\cite{devlin2018bert} or
GPT~\cite{gpt-2018,gpt2-2019,gpt3-2020} constantly advance the state of the art
on many natural language processing tasks, it is interesting to evaluate them on
more specialized tasks.
Multiple surveys~\cite{minaee2024llmsurvey,zhao2023survey} and
benchmarks~\cite{papcode2024hellaswag,chiang2024chatbot} show that large
language models are good at following instructions.
We can also intuit that training LLMs on diverse data (for instance the
Pile~\cite{gao2020pile}) uniquely qualifies them to produce text in a wide
variety of styles.

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended to entire articles or corpuses of text because
these are also the object of linguistic style and
stylistics~\cite{lugea2023stylistics}.
The task of transferring text style has a certain maturity.
The interest in this task is renewed by the advancements made with LLMs.

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
More often than not, these values are easy to compute while the methods used for
obtaining them are usually idempotent.
This makes them very attractive because of their dual potential purpose.
On one hand, they can be used to instruct the generation of text (``write no
more than 10 sentences'').
On the other hand they can be used to analyse text.
This ability allows verifying whether a text has the expected value for an HLF.

Taking Noam Chomsky's ideas~\cite{chomsky2002syntactic} that syntax and meaning
are different things, we intuit that we can define writing style only in terms
of both syntax and meaning.
We shy away from discussing meaning, but syntax and orderly constructs like it
seem to be within our grasp.

Traditionally, linguistic style was thought to be the result of morphological
and syntactic arrangements~\cite{lugea2023stylistics}.
More recent schools of thought on linguistic style find semantics and pragmatics
to be inextricably linked to author style~\cite{verma2019lexical}, but argue
that a great deal can still be inferred from lexicon, syntax and a `surface'
level which contains basic HLFs such as the number of words in a sentence.

Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

These are the strands of thought that have inspired us to pursue two research
questions in the age of LLMs.

@@

Linguistic style is the sum of identifiable language choices manifest in a text
made from the language system by the text producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
The ample literature on the subject seems to agree that style depends on context
and author choice vis-a-vis a communication
goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
This work grounds our choice of handcrafted linguistic features related to
linguistic style.

@@

Our paper describes an experiment in controllable text generation
(CTG)~\cite{zhang-ctg-2022}.
Hightened recent interest in CTG and especially in benchmarking LLMs in the
context of CTG~\cite{chen2024benchmarking} is particularly relevant for this
paper.
Such research alleviates the burden of proof regarding the level to which
LLMs respond well to diversified instructions for controlled text generation in
general.
This paper differs from the existing CTG work with its use of HLFs for writing
the text generation instructions.
It additionally uses HLFs to measure the performance of LLMs in terms of their
ability to generate text that conforms to some chosen target HLF values.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.

To run the experiment we must select the LLMs to evaluate and finding datasets
from which we can derive corpuses.
Then there's the matter of selecting the HLFs for our experiment.

We think it's best to describe the experimental process before going over our
selection for the experiment's parameters.

\subsection{Process}\label{subsec:experiment-process}

Assume we have a target LLM, a corpus containing texts, some examples extracted
from the corpus, a selection of HLFs and input text both from within and from
outside the corpus.
This description of the experiment process assumes those parameters have stable
values.
Changing the parameters requires restarting the process.

The first activity involves establishing baselines for each experiment scenario.

In the first scenario the target LLM is instructed to generate text retaining
the meaning of  the input text without any further instruction related to style
or the linguistic features of the generated text.
This step is repeated 10 times, computing each HLF in our selection for the
generated text.
The computed values make up the first scenario baselines for each HLF.

The second scenario consists in generating text from the same input text, but
with the aid of some examples from the corpus.
In this scenario, we believe that the choice of input text affects the outcome.
Therefore we propose two variants for this scenario: one with input text
extraneous to the corpus and one with input text from the corpus.

We add the examples to the initial LLM prompt and follow the same process like
in the first scenario.
We do this for each variant of the second scenario to obtain two more baselines
for each HLF for the corpus.

The number of examples is chosen so that we do not overrun the context window of
the LLM.
Some context windows are smaller than others, settling the number of examples at
three.

The next step requires computing statistics on the corpus for each HLF:
\begin{itemize}
    \item \textbf{min}: the minimum value of the HLF on the corpus;
    \item \textbf{max}: the maximum value of the HLF on the corpus;
    \item \textbf{avg}: the mean value of the HLF on the corpus.
\end{itemize}

Doing so allows writing an HLF prompt derived from each of our baseline prompts.
The HLF prompts instruct the LLM to generate text whose HLFs comply with the
previously computed statistics.
The individual (per-HLF) instructions are all are part of the same HLF prompt.
The prompts are constructed using Jinja templates which are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/src/llm_tst_consistency/templates}{templates}}
directory of the source code used to run the experiment.
The instruction prompts corresponding to our two scenarios are
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The rest of the templates in the directory are used to construct the individual
HLF instructions.

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.

The first scenario is indicative of whether LLMs acknowledge HLF instructions.
If they do then the results obtained from HLF instructions should be
significantly different from the corresponding baseline results.

Both variants of the second scenario quantify how similar the HLFs of the texts
generated by the LLM are to the corpus average \textit{because of} the HLF
instructions.
The rationale here is that the LLM can generate text very similar to text from
the corpus by using the supplied examples from the corpus.
That would leave the HLF instructions as the remaining possible cause for
generating text with HLFs that are consistently different from the corresponding
baseline.

@@

We only use state of the art large language models for the experiment.
The Chatbot Arena~\cite{chiang2024chatbot} leaderboard is essential in our
selection.
To cover more ground we find top performers from different vendors interesting.
The final selection is available in Table~\ref{table-llm}.

@@

The ID of the LLM is referenced throughout figures, tables and generally
throughout the rest of the paper.

@@

The instructions for each HLF are specifically crafted using natural language.
The resulting instructions can be catalogued on the concrete-abstract spectrum.
For example, `write 100 words' is a more concrete instruction than `write as if
you were in junior highschool'.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category.
Table~\ref{table-hlf} showcases our selection of features.

@@

The LFTK ID is used throughout tables and figures to refer to a specific HLF.

\subsection{Corpus Selection}\label{ds-selection}

The experiment design rests upon measuring HLFs on a corpus of texts and
computing statistics.
LLMs also make use of examples extracted from a corpus to guide text generation.
Our approach idempotently derives a corpus of texts from a well known dataset.
Our selection of datasets is listed in Table~\ref{table-ds}.

@@

The Yelp review data set provides a diverse assortment of reviews from online
users.
The authors employ a casual writing style, very similar to the writing style of
an average person.
We construct a corpus containing the first 1000 reviews from the test split of
this dataset.

CNN/DailyMail contains stories and news with diverse styles.
The style varies from one author to another and it also follows the publisher's
editorial style.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

@@

This experiment presupposes instructing various large language models to reword
an input text so that it conforms to some HLFs.

There is an input text corresponding to each corpus that is external to that
corpus.
These texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.

For the variant that uses an input text from the corpus, we draw the text
randomly from the corpus once, at the start of the experiment.

The input text that is external to the CNN article corpus is an extract from the
classic \textit{The Life and Opinions of Tristram Shandy, Gentleman}, by
Laurence Sterne~\cite{sterne2003life}.
The text style is very dissimilar compared to news articles.
Its linguistic features are also very unlike the average on the CNN article
corpus.
A text that is similarly divorced from the style and linguistic features of most
Yelp reviews is Oscar Wilde's \textit{Sonnet toLiberty}~\cite{wilde1909poems}.

Choosing input texts that are vastly different from their corresponding corpus
is based in the assumption that LLMs would generate baseline results that do not
conform to the HLF statistics on that corpus.

@@

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
Plot figures show the HLF on the vertical axis, while the horizontal axis is
reserved for the trial number.
HLF values for the text generated using baseline prompts are plotted as a
continuous grey line, while the values obtained by using HLF enhanced prompts
are plotted using a continuous grenat line.
The minimum and maximum HLF on the benchmark dataset are plotted using dotted
black lines.
A continuous black line designates the average HLF on the corpus.
We sometimes refer to this average as the \textit{target} value because the LLM
is instructed to generate text that primarily conforms to this value.

Result tables highlight the difference between plots obtained using a
baseline prompt versus the ones obtained using HLF instruction prompts.
Firstly, we measure how different baselines are from HLF-enhanced plots.
Secondly, we compare the closeness of the baseline and HLF-enhanced plots to the
target average.
Table values where the text is \textbf{bold} mark significant differences
between baseline and HLF instruction results.
\underline{Underlined} values mark situations where the result of using HLF
instructions is closer to the target than the baseline.
\textit{Italic} values mark situations where the baseline is closer to the
target.

\subsection{Measurements}\label{subsec:measurements}

@@

To determine which of two results is closer to the target, we compute the area
between each result and the target.
Another way in which we measure the closeness of the result to the target is by
computing the Euclidian norm of the vectorial difference between the result and
target.

Note that each HLF can have a different range of values than other HLFs.
When we compute the area and the Euclidian norm, we use the raw values of each
HLF, without preprocessing them.
Therefore the results for one HLF are not comparable with the results for
another HLF.

Remember that we don't quantify how different the baseline is from the
HLF-enhanced result.
Instead we calculate the relative closeness to a common target.
Because we took this approach, comparing values for the same HLF, but different
LLMs is also ill-advised.

Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
instruction result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
We define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$.

@@

In the next sections we present our perspective on the most relevant results of
the experiment.

@@

The results on the CNN article corpus showcased in Table~\ref{table-prompt-1-cnn-dailymail}
shows that LLMs take into account instructions concerning the HLFs of their
output.

@@

In the second scenario of the experiment we try to quantify the impact of using
HLF instructions on the output of the LLM.
There are two variants of this scenario: input text from within or from outside
the corpus.

Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant.
The outlook is similar to the first experiment scenario.

@@

A possible explanation for having fewer significant differences in results might
lie in the closeness to the target.
As we can see in Figure~\ref{fig-p2-gpt-twords}, this is not always the case.
In~\cref{fig-p2-command-r-simpttr,fig-p2-gemini-a-kup-pw} we observe regressions
in performance with significant differences between baseline and result.

@@

Even so, the results obtained in our first scenario are largely confirmed in the
first variant of our second scenario.

We finally look at the second variant of the second scenario of our experiment.
The hypothesis in this variant is that LLMs will generate text that exhibits
HLFs much closer to the target corpus average if the input text is from that
corpus.
The expectation is to see less significant results.

@@

The results shown in Table~\ref{table-prompt-2-ifd-cnn-dailymail} contradict our
expectations outright.
The most surprising behaviour is exhibited for HLFs which are derived into more
concrete generation instructions, such as the total number of words.
This behaviour is highlighted in figure~\ref{fig-p2-ifd-gemini-twords}.
Additionally, we notice regressive behaviour for some models compared to the
previously explored results in figure~\ref{fig-p2-ifd-claude3-nuadj}.
Finally, the abstractness of the text generation instructions for certain HLFs
leads to the generated text not being compliant as exemplified in
figure~\ref{fig-p2-ifd-llama3-a-kup-pw}.

@@

The results in this last scenario show that whether LLMs take examples listed in
the prompt into account is debatable when it comes to generating text with
similar linguistic features as the examples'.

@@

In terms of our first question --- whether LLMs are able to follow instructions
concerning handcrafted linguistic features --- the response is in our opinion
affirmative.
Certainly, not all LLMs are created equal in this regard.

The design of the experiment leaves room for improvement because of the choices
regarding computing HLF statistics on a corpus, input text, the corpus
itself or the instruction prompt.
The importance of these factors is evidentiated by the difference in results
obtained using the Yelp reviews corpus when compared to the results obtained
on the CNN article corpus.
While these results also show a significant influence when using HLFs in the
prompt, the influence is mostly negative.

This leads us to our second question: how much do handcrafted linguistic feature
instructions affect the way LLMs write text.
The results in this regards show us a much less conclusive picture.
It is not apparent the LLM will follow HLF instructions as intended.
In fact, using the same prompt on different data lead to wildly different
outcomes.
Whether this is because of circumstances such as the selection of HLFs, the LLM,
the HLF target values, the input text, a.s.o remains to be determined.
We can observe some idioms, though.

The more abstract the instruction to comply to an HLF, the fewer the chances of
a significant impact on the outcome.
When the outcome is significant, it tends to be negative.
This tells us that large language models are likely good at following
instructions about HLFs when those instructions are more concrete.

Then, the choice of input text in relation to the target HLF values seems very
important.
LLMs don't seem able yet to cover the vast gulf between the linguistic features
of an Oscar Wilde poem and the average linguistic features of text written by
Yelp reviewers.
This leads us to propose that even if LLMs are able to generate text that
complies to certain target HLFs, there is a limit to this ability.

Finally, the overall selection of HLFs is important.
Choosing HLFs that measure the uniqueness of a linguistic element (words,
structures, a.s.o) limits the LLMs ability to conform to other HLFs (for example
total number of sentences or reading ease).
This is highlighted in the experiment specifically in the tension between the
age of acquisition and the adjective or verb variation.

@@

We started out by trying to understand whether it is possible to generate text
that exhibits certain linguistic features using a large language model.
To our surprise, it turns out that instructing language models to write text
that conforms to a selected assortment of linguistic features almost always
leads to significantly different linguistic featues of generated text.

The results are not always satisfying, though.
At this stage, the more pragmatic answer seems to involve prompt engineering and
a careful choice of language features.
Providing examples does not seem to influence the model's ability to comply with
the user's desires.

Other experiments are required to better understand the capabilities of large
language models with regard to handcrafted linguistic features.
For example, testing the response of language models to features in isolation
or not using a corpus or input text are different setups that could eliminate
some of the unknowns left uncovered by the current experiment.
